---
title: Homework 1
---
The purpose of this document is to get you some basic practice using *R 
  Markdown*. As mentioned in the syllabus, I will also be asking for all 
  homeworks to be completed with *R Markdown* so it's worth getting used to now.

To receive full credit, please create and render an *R Markdown*  document in 
  html or pdf format that has the following elements.
    - Title, author and date
    	+ Use a YAML header in your document to do this
    - An R code chunk displaying how to load data into R and store it into 
    	an object, along with text explaing in the code.
    - A generic ordered or unordered list with at least one level of nesting 
      (could even be a shopping list)
    - Some R output of some sort
    - Calculate the mean of a variable and display it in text. Make some
    	manipulation to the data (e.g., remove a few random cases) and 
    	report the mean in text again, using code. For 
    	example:

```{r}
data(ChickWeight)
head(ChickWeight)
```

```{r, echo = FALSE}
options(digits = 2)
```
*The average chick weight was `r mean(ChickWeight$weight, na.rm = TRUE)` grams. 
  At time time point 0 the average weight was 
  `r tapply(ChickWeight$weight, ChickWeight$Time, mean, na.rm = TRUE)[1]` 
  grams.*

To see the datasets available in R, type in `data()` at the command prompt. You
  should see a summary of many different datasets. To get information on any one
  of these datasets, type `?` and the name of the dataset. For example 
  `?ChickWeight` will tell you some information about the dataset above.

## Plot
For the final portion of this homework, please pick a dataset (please pick one
  other than the ChickWeight dataset) and produce a `pairs()` plot. I encourage 
  you do play around with this and check out the documentation. Basically, you 
  just need to feed it a dataframe with the variables you want to visualize 
  through the scatterplot matrix. For example

```{r}
pairs(ChickWeight)
```


Now you just need to render the document using either the `knit` function in 
  RStudio, or using `knit::knit2html("yourFile.Rmd")`.