---
title: Homework 1
output: pdf_document
---
The purpose of this document is to get you some basic practice using *R 
  Markdown*, as well as to get some practice reading in data and working with
  vectors and matrices. As mentioned in the syllabus, I will be asking for all 
  homeworks to be completed with *R Markdown* so it's worth getting used to now.

To receive full credit, please create and render an *R Markdown*  document in 
  html or pdf format that has the following elements.

- Title, author and date
    + Use a YAML header in your document to do this
- An R code chunk displaying how to load data into R and store it into an 
  object, along with text explaining  the code.
- A generic ordered or unordered list with at least one level of nesting (could 
  even be a shopping list)
- Some R output of some sort
- Calculate the mean of a variable, via `mean()` and display it in text. If 
  there are missing data in the vector that you are trying to calculate the mean
  from, you will also need to include the additional argument `na.rm = TRUE`. 
  Make some manipulation to the data (e.g., remove a few random cases) and 
  report the mean in text again, using code. For example:

```{r}
data(ChickWeight)
head(ChickWeight)
# globally round all output to two digits
options(digits = 2)
```

*The average chick weight was `r mean(ChickWeight$weight, na.rm = TRUE)` grams. 
  At time point 0 the average weight was 
  `r tapply(ChickWeight$weight, ChickWeight$Time, mean, na.rm = TRUE)[1]` 
  grams.*

## Plot
Use the dataset you loaded, or pick a new dataset from base R to produce a 
  `pairs()` plot (scatterplot matrix). To see the datasets available in R, type 
  `data()` at the command prompt. You should see a summary of many different 
  datasets. To get information on any one of these datasets, type `?` and the 
  name of the dataset. For example `?ChickWeight` will tell you some information 
  about the dataset above. If you choose to use a base R dataset, please use one 
  other than the ChickWeight dataset.

  To produce a pairs plot, you just need to feed it a data frame with the 
  variables you want plotted. I encourage you to play around with this function 
  and check out the documentation. For example, to get a scatterplot matrix of 
  all but the third variable in the ChickWeight dataset, use

```{r}
pairs(ChickWeight[ ,-3])
```

## Matrices and Vectors
Create the following matrix (which we created during the first class, if you
  saved your syntax)
$$\mathbf{m} = 
\left [
  \begin{array}{ccccc}
    18 & 32 & 11 & 41 & 73 \\\
    61 & 47 & 22 & 87 & 63 \\\
    44 & 52 & 23 & 42 & 77 \\\
    23 & 17 & 5 & 72 & 83
  \end{array}
  \right ]
$$

1) Use three different methods to subset the matrix and obtain the following 
  vector
$$\mathbf{v} = 
\left [
  \begin{array}{ccc}
    52 & 42 & 77
  \end{array}
  \right ]
$$

2) Subset **m** again to get 
  $\mathbf{v2} = \left [\begin{array}{cc} 87 & 42 \end{array} \right ]$. Then 
  create a new matrix using the following steps:
     + Drop the final column of matrix **m**
     + Add **v2** to the second and fourth rows (this will utilize the property 
       of recycling). 


## Render the document
Once you have fulfilled all the requirements of the homework, you  need to 
  render it to html (recommended) or pdf (requires additional software 
  installations) using either the `knit` function from the `knitr` package, or 
  `render` from the `rmarkdown` package. The `knit2html` button in Rstudio will 
  essentially just run the code for you.

If you're using RStudio, you shouldn't have to worry about much to get it 
  rendered. If not, make sure you have the correct packages installed. The code
  below should help.

```{r, eval = FALSE}
#### Via the knitr package
# install package
install.packages("knitr") # note, this only needs to be done once
library(kntir)
knit2html("yourFile.Rmd")

#### Via the RMarkdown package
install.packages("rmarkdown")
library(rmarkdown)
render("yourFile.Rmd", "html_document")
```