---
title       : "String Manipulations: grep and regex"
subtitle    : Week 7.2
author      : Daniel Anderson
job         : CourseR
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
--- &twocol
```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = "/Users/Daniel/Dropbox/Teaching/CourseR/")
```
<style>
em {
  font-style: italic
}
</style>

<style>
strong {
  font-weight: bold;
}
</style>

## Today's Agenda
String Searching and regular expressions

* Searching strings
	+ `grep()`
	+ `grepl()`
	+ `sub()`
	+ `gsub()`
* Regular expressions
	+ metacharacters
	+ character classes

----
## Needle in a haystack
`grep` and `grepl` can be used to find a needle (string pattern) in a haystack string vector. By default, `grep` will return the location in the vector of any string matching the pattern.

```{r grep1}
fruits <- c("apple", "orange", "banana", "mango")
grep("b", fruits)
grep("o", fruits)
```

-----
## Return values, rather than an index

```{r grep2}
grep("b", fruits, value = TRUE)
grep("o", fruits, value = TRUE)
```

----
## Logical version: `grepl`

```{r grepl1}
grepl("b", fruits)
grepl("o", fruits)

```

----
## Using `grep` and `grepl` for subsetting

```{r loadScCard}
# load data
scCard <- read.csv("./data/CollegeScorecard.csv", 
	stringsAsFactors = FALSE, na = c("NULL", "PrivacySuppressed"))
head(scCard)
```

----
## Select Institution ID and SAT variables

```{r selectSATvars}
sat <- scCard[ ,c(1, grep("SAT", names(scCard)))]
head(sat)
```

---- &twocol
## with `grepl`
# Two steps, two methods

*** =left

```{r greplSATvars1}
satVars <- grepl("SAT", names(scCard))
satVars[1] <- TRUE
sat <- scCard[ ,satVars]
head(sat)
```

*** =right

```{r greplSATvars2}
satVars <- scCard[ ,grepl("SAT", names(scCard))]
sat <- as.data.frame(
		cbind("UNITID" = scCard[ ,1], satVars))
head(sat)
```

----
## Percentage degrees awarded by program type

<span style="color:gray"> see full data dictionary at </span> https://collegescorecard.ed.gov/data/documentation/

```{r scCardDict, echo = FALSE, results = "asis"}
vars <- c("PCIP01", "PCIP03", "PCIP04", "PCIP05", "PCIP09", "PCIP10", 
			"PCIP11", "PCIP12", "PCIP13", "PCIP14", "PCIP15", "PCIP16", 
			"PCIP19", "PCIP22", "PCIP23", "PCIP24", "PCIP25", "PCIP26", 
			"PCIP27", "PCIP29", "PCIP30", "PCIP31", "PCIP38", "PCIP39", 
			"PCIP40", "PCIP41", "PCIP42", "PCIP43", "PCIP44", "PCIP45", 
			"PCIP46", "PCIP47", "PCIP48", "PCIP49", "PCIP50", "PCIP51", 
			"PCIP52", "PCIP54")
desc <- c("Agriculture, Agriculture Operations, and Related Sciences", 
			"Natural Resources and Conservation", 
			"Architecture and Related Services", 
			"Area, Ethnic, Cultural, Gender, and Group Studies", 
			"Communication, Journalism, and Related Programs", 
			"Communications Technologies/Technicians and Support Services", 
			"Computer and Information Sciences and Support Services", 
			"Personal and Culinary Services", "Education", "Engineering", 
			"Engineering Technologies and Engineering-Related Fields", 
			"Foreign Languages, Literatures, and Linguistics", 
			"Family and Consumer Sciences/Human Sciences", 
			"Legal Professions and Studies", 
			"English Language and Literature/Letters", 
			"Liberal Arts and Sciences, General Studies and Humanities", 
			"Library Science", "Biological and Biomedical Sciences", 
			"Mathematics and Statistics", 
			"Military Technologies and Applied Sciences", 
			"Multi/Interdisciplinary Studies", 
			"Parks, Recreation, Leisure, and Fitness Studies", 
			"Philosophy and Religious Studies", 
			"Theology and Religious Vocations", "Physical Sciences", 
			"Science Technologies/Technicians", "Psychology", 
			"Homeland Security, Law Enforcement, Firefighting and Related Protective Services", 
			"Public Administration and Social Service Professions", 
			"Social Sciences", "Construction Trades", 
			"Mechanic and Repair Technologies/Technicians", 
			"Precision Production", "Transportation and Materials Moving", 
			"Visual and Performing Arts", 
			"Health Professions and Related Programs", 
			"Business, Management, Marketing, and Related Support Services", 
			"History")
knitr::kable(data.frame(vars, desc), 
	col.names = c("Variable Name", "Program"))
```

----
## Select percentage degree awarded var

```{r pctSelect}
pct <- scCard[ ,c(1, grep("PCIP", names(scCard)))]
head(pct)
```

----
## Create a new max variable
Variable should include the **name** of the PCIP that corresponds to the institution's maximum.

```{r maxProp}
# Remove rows with complete missing data
pct2 <- pct[apply(pct[ ,-1], 1, function(x) sum(is.na(x)) != (ncol(pct) -1)), ]

maxLoc <- apply(pct2[ ,-1], 1, which.max) + 1
pct2$maxprop <- sapply(1:nrow(pct2), function(i) {
	names(pct2)[maxLoc[i]]
})
head(pct2[ ,35:40])
```

----
## Merge results back into full dataset, plot results

```{r mergeMaxProp1, eval = FALSE}
scCard <- merge(scCard, pct2[ ,c(1, ncol(pct2))], by = 1, all = TRUE)
par(las = 2)
plot(table(scCard$maxprop),
	main = "Frequency of Max Graduation Programs",
	xlab = "Program Code",
	ylab = "Frequency",
	tck = 0,
	cex.axis = 0.6)
```

----

```{r mergeMaxProp, fig.height = 9, fig.width = 14, echo = FALSE}
scCard <- merge(scCard, pct2[ ,c(1, ncol(pct2))], by = 1, all = TRUE)
par(las = 2)
plot(table(scCard$maxprop),
	main = "Frequency of Max Graduation Programs",
	xlab = "Program Code",
	ylab = "Frequency",
	tck = 0,
	cex.axis = 0.6)
```

----
## Selecting rows
Select only Oregon schools

```{r rowSelect1}
OR <- scCard[grep("OR", scCard$STABBR), ] # grepl could be used equivalntly
table(OR$STABBR)
head(OR)
```

----
## Selecting rows (continued)
Select only schools with ".edu" in their website (i.e., eliminate hair schools, etc.)

```{r rowSelect2}
eduWeb <- scCard[grep(".edu", scCard$INSTURL), ]
head(eduWeb)
```

----
## Select only community colleges

```{r rowSelect3}
Sys.setlocale('LC_ALL','C')
# I was getting some weird warnings with the code below. Google 
# helped me fix it with the above line of code. Still not sure 
# what was going on.

cc <- scCard[grep("Community", scCard$INSTNM), ]
head(cc)
```


----
## Substituting characters

```{r sub1}
sentences <- c("Here's some text. Here's some more text.", 
			  "Red text makes me feel bad.",
			  "Green text makes me feel good.")
sub("text", "food", sentences) # replace only the first match within each element
gsub("text", "food", sentences) # replace all matches
```

----
## Real example

```{r sub2}
cam <- read.csv("./data/CamdenBoroughs.csv", na = "999")
head(cam$Percentage)
cam$frl <- as.numeric( sub("%", "", cam$Percentage.Claiming.Free.School.Meals) )
str(cam$frl)
```
Note that in the above we can use either `sub` or `gsub` equivalently, because there's only one instance. In other cases, this would not be the case. For example, if we wanted to change the format of the dates.

-----
## Change census date format

```{r censusDateReFormat}
cam$censusDate <- sub("/", "-", cam$Census.Date) 
head(cam$censusDate) # Not what we want
cam$censusDate <- gsub("/", "-", cam$Census.Date) 
head(cam$censusDate) # What we want
```

---- .segue
# Regular Expressions

----
## What is a regular expression?
* "an 'instruction' given to a function on what and how to match or replace strings" (Eden, 2007)
* *Metacharacters* are special characters that define specific operations
	+ can be interpreted as standard characters, provided the appropriate syntax is used.
	+ include the following: 

```{r regexpTbl, echo = FALSE}
noquote(c("$","*","+",".","?","[","^","{","|","(","\\"))
```
* *Sequences* define sequences of characters to match
* *Character classes* define ranges to match or not match

----
## Why does this matter?
Many of the built-in string functions in R take regular expressions as their arguments. If you're unaware of how regular expressions work, you could end up with unexpected behavior. For example, the following seems like it should work, but it will not.

```{r strSplitRegEx1}
string <- "School is fun. Especially recess. That's the best part. I love recess."
strsplit(string, ".")
```

----
We can get the behavior we intend by overriding the metacharacter

```{r strSplitRegEx2}
strsplit(string, "\\.")
```

----
## Quick overview of metacharacters
(we'll talk about each in more detail)

```{r metaTable, echo = FALSE, results = "asis"}
sequences <- noquote(c("$","*","+",".","?","[","^","{","|","(", "\\"))
operation <- c("Matches the end of a string",
			   "Matches preceding character at least 0 times",
			   "Matches preceding character at least 1 time",
			   "Matches any single character (i.e., skip operator)",
			   "Matches preceding character 0 or 1 time",
			   "Define character list or character classes",
			   "Matches the start of a string",
			   "Define n to m matches of preceding character",
			   "Or operator",
			   "Define groupings for backreferencing",
			   "Suppress metacharacters and define anchor sequences")
knitr::kable(data.frame("Metacharacter" = sequences, 
						"Operation" = operation))
```

----
## Quantifiers
```{r quantifiers, echo = FALSE, results = "asis"}
knitr::kable(data.frame("Metacharacter" = sequences, 
						"Operation" = operation)[c(2, 3, 5, 8), ], 
	row.names = FALSE)
```
<span style="color:gray; font-size:8pt;">Examples taken from http://stat545.com/block022_regular-expression.html </span>

```{r quantifiers1}
letterSet <- c("a", "ab", "acb", "accb", "acccb", "accccb")
grep("ac*b", letterSet, value = TRUE)
grep("ac+b", letterSet, value = TRUE)
```

----
```{r pTable1, echo = FALSE, results = "asis"}
knitr::kable(data.frame("Metacharacter" = sequences, 
						"Operation" = operation)[c(5, 8), ], 
	row.names = FALSE)
```

```{r pLetterSet1, echo = FALSE}
letterSet

```

```{r quantifiers2}
grep("ac?b", letterSet, value = TRUE)
grep("ac{2}b", letterSet, value = TRUE)
grep("ac{2,}b", letterSet, value = TRUE)
```

-----

```{r pLetterSet2, echo = FALSE}
letterSet
```

```{r quantifiers3}
grep("ac{1,3}b", letterSet, value = TRUE)
grep("ac{0,3}b", letterSet, value = TRUE)
```

---- &twocol
## Position matching

*** =left

```{r posMatchTab1, results = "asis", echo = FALSE}
knitr::kable(data.frame("Metacharacter" = sequences, 
						"Operation" = operation)[c(1, 7, 11), ], 
	row.names = FALSE)
```
Anchor sequences are provided on the right (from Sanchez, p. 61). These should not be confused with R escape characters, such as `\n` and `\t`, for new line and tab, respectively. 

*** =right

<div align = "center">
<img src = ./assets/img/anchorSequences.png width = 500 height = 500>
</div>

----
## Match the end of a string

```{r posMatch1}
itemIDs <- c("RF3L02E03", "RF3M08E05", "RF3H10E08", "RL1L03E03", "RL1M05E05",
 			 "RL1H10E04", "RI2L03E07", "RI2M06E05", "RI2HSAMPLEE06", 
 			 "WR4L02E03", "WR4M06E06", "WR4H09E03", "WR9L03E04", "WR9M08E04", 
 			 "WR9H12E04", "LA1L01E11", "LA1M06E04", "LA1H09E04", "WR2L03E05", 
 			 "WR2M06E05", "WR2H10E05", "LA2LSAMPLEE03", "LA2M06E04", 
 			 "LA2H09E08", "RF4L02E03", "RF4M08E06", "RF4H09E07", "RL7L02E07", 
 			 "RL7M06E06", "RL7H10E06", "RI1L02E07", "RI1M07E08", "RI1H11E07", 
 			 "WR1L02E07", "WR1M07E07", "WR1H11E08")

# Select fourth grade items
grep("4$", itemIDs, value = TRUE)
```

----
## Match the start of a string

```{r posMatch2}
# Select RL items
grep("^RF", itemIDs, value = TRUE)

# Select WR items
grep("^WR", itemIDs, value = TRUE)
```
<br>

Note that in this case, the result would be the same without using the `^` metacharacter, but it's safer to go ahead and use it anyway. In other cases, there may be matches with the same pattern that are not at the beginning.

----
## Other examples

<span style="color:gray; font-size:8pt;">Examples taken from http://stat545.com/block022_regular-expression.html </span>

```{r posMatch3}
string2 <- c("abcd", "cdab", "cabd", "c abd")
grep("ab", string2, value = TRUE)
grep("^ab", string2, value = TRUE)
grep("ab$", string2, value = TRUE)
```

----
## Search start of words, rather than strings

```{r printString2, echo = FALSE}
string2
```

```{r posMatch4}
grep("\\ba", string2, value = TRUE)
grep("\\b a", string2, value = TRUE)
```

----

```{r posMatch5}
ids <- c("123225-5417", "132123-132975", "321579-123569")
grep("123", ids, value = TRUE)
grep("^123", ids, value = TRUE)
grep("\\b123", ids, value = TRUE) # New word starts with 123
grep("\\B123", ids, value = TRUE) # New word does not start with 123
```

----
## Digits

```{r digits}
digits <- c("Charlie", "Charlie2", "Mary", "Marianne", "Mary2", "15")
grep("\\d", digits, value = TRUE)
grep("\\D", digits, value = TRUE)
```

----
## Spaces

```{r spaces}
string
gsub("\\s", "_", string)
gsub("\\S", "_", string)
```

----
## Words

```{r words}
gsub("\\w", "_", string)
gsub("\\W", "_", string)
```

----
## Operators

```{r operatorTab1, results = "asis", echo = FALSE}
knitr::kable(data.frame("Metacharacter" = sequences, 
						"Operation" = operation)[c(4, 6, 9, 10, 11), ], 
	row.names = FALSE)
```

-----
## Skip characters

```{r skip1}
string2 
grep(".ab", string2, value = TRUE)
grep("ab.", string2, value = TRUE)
```

----
## Select high items (easy)
```{r skip2}
head(itemIDs)
grep("H", itemIDs, value = TRUE)
```

----
## Select low items (more difficult)

```{r skip3}
head( grep("L", itemIDs, value = TRUE) ) # Fails
grep("...L", itemIDs, value = TRUE) # Fails
grep("^.{3}L", itemIDs, value = TRUE) # Success!
```

----
## Or

Select RL or WR Items

```{r or1}
grep("RL|WR", itemIDs, value = TRUE)

```

----

```{r or2}
string3 <- c("^ab", "ab", "abc", "abd", "abe", "ab 12")
grep("bc|be", string3, value = TRUE)
grep("1|c", string3, value = TRUE)
```


----
## Backreferences

Find and reference any number of patterns.

```{r backreferences1}
catMat <- "The cat sat on the mat."
gsub("(.at)", "\\1!", catMat)
gsub("(.)at", "\\1ot", catMat)
gsub("(.he) (.at)", "__\\1__ \\2!", catMat)
```
<span style="color:gray; font-size:8pt;">Examples adapted from https://www.stat.auckland.ac.nz/~paul/ItDT/HTML/node84.html </span>


----
## Suppress metacharacters

```{r suppress1}
string3
grep("^ab", string3, value = TRUE)
grep("\\^ab", string3, value = TRUE)
```

----

```{r suppress2}
string
gsub(".", "___", string)
gsub("\\.", "____", string)
```

----
## More examples
```{r regEx+Example}
set.seed(300)
d <- data.frame(rnorm(3), rbinom(3, 1, 0.5), rpois(3, 2))
d
names(d) <- gsub(pattern = "\\.+", replacement = "", names(d))
```

---- .segue
# Character classes

----
## Search/specify entire classes of characters

```{r charLists1}
string3 <- c("^ab", "ab", "abc", "abd", "abe", "ab 12")
grep("ab[cd]", string3, value = TRUE)
grep("ab[c-e]", string3, value = TRUE)
```

----
```{r charLists2}
textNum <- paste0(rep(letters[1:14], each = 7), 1:7)
textNum
grep("[a-e][3-5]", textNum, value = TRUE)
```
Invert the character list with `^`.

```{r charLists3}
grep("[^a-e][3-5]", textNum, value = TRUE)
```

----
## POSIX character classes

```{r charClassTab, results = "asis", echo = FALSE}
classes <- c("[:digit:]", "[:lower:]", "[:upper:]", "[:alpha:]", "[:alnum:]",
			 "[:xdigit:]", "[:blank:]", "[:space:]", "[:punct:]", "[:graph:]",
			 "[:print:]", "[:cntrl:]")
classDesc <- c("Digits, 0-9", "Lower case letters", "Upper case letters",
			   "Alphabetic characters", "Alphanumeric characters", 
			   "Base 16 hexadecimal", "Blank characters (space and tab)", 
			   "Space characters (all)", "Punctuation characters",
			   "Human readable characters", "Printable characters", 
			   "Control characters (line breaks, etc.)")

equivalence <- c("[0-9]", "[a-z]","[A-Z]", "[[:lower:][:upper:]] or [A-z]",
				"[[:alpha:][:digit:]] or [A-z0-9]", "[0-9A-Fa-f]", "", "", "",
				"[[:alnum:][:punct:]]", "[[:alnum:][:punct:]\\\\\\s]", "")
knitr::kable(data.frame(classes, classDesc, equivalence), 
	col.names = c("Code", "Description", "Equivalent Code"),
	caption = "Character Classes")
```

----

```{r charClass1}
string4 <- c("ab", "ab12", "Dc a", "BB", "Here's some text.", "Here's a \n new line.")
cat(string4)
grep("[[:digit:]]", string4, value = TRUE)
grep("[[:lower:]]", string4, value = TRUE)
grep("[[:upper:]]", string4, value = TRUE)
```

----
```{r charClass2}
grep("[[:space:]]", string4, value = TRUE)
grep("[[:punct:]]", string4, value = TRUE)
strsplit(sentences, "[[:space:]]")
```

----
```{r charClass3}
strsplit(sentences, "[[:space:][:punct:]]")
```

-----
## Things we still haven't covered
* `sprintf` (useful for padding numbers)
* `regexpr`
* `gregexpr`
* `regexec`
* Matching
	+ `regmatches`
	+ `match`
	+ `pmatch`
	+ `charmatch`
* (...still more I'm probably missing)
* *stringr* package (part of the Hadleyverse, contains many useful functions for manipulating strings)

---- .segue
# Example with real data
<span style="color:gray; font-size:8pt;">Examples taken from https://rstudio-pubs-static.s3.amazonaws.com/74603_76cd14d5983f47408fdf0b323550b846.html </span>

----
## Gapminder data

```{r gapMinder1}
install.packages("gapminder")
library(gapminder)
data(gapminder)
head(gapminder)
```

-----
## Quick aside
Data are also available online. We haven't talked about this much, but you can always read data into R directly from the web

```{r webGapMinder}
gapMinder <- read.delim("http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt", 
	header = TRUE)
head(gapMinder)
```

----
## Searching countries
Find all countries that end with *land*, and have an "i" or "t" in them.

```{r gapMinder2}
grep("(.*[it].*)land$", gapminder$country, value = TRUE)
```

----
## Backreferencing
Replace *land* for *LAND* for countries with an "i" or "t" in them

```{r gapMinder3}
gapminder$country <- gsub("(.*[it].*)land$", "\\1LAND", gapminder$country)
unique(
	grep("LAND", gapminder$country, value = TRUE)
	)
```

----
## Abbreviated lab
* Find countries that begin with "South"
* Find countries that have "Ga" at the beginning of either word (not string)
* Find countries that have 2 consecutive e's

# Advanced
* write a function that takes a generic string variable and makes the following manipulations: 
	+ keeps only alpha-numeric characters
	+ removes all spaces
	+ converts to lower case
As an example, if the string Coeur d'Alene were fed to the function, it should return coeurdalene. (you can use this to test your function)

<span style="color:gray; font-size:8pt;">Examples taken from https://rstudio-pubs-static.s3.amazonaws.com/74603_76cd14d5983f47408fdf0b323550b846.html </span>

----
# Find countries that begin with "South"

```{r lab1}
unique( grep("^South", gapminder$country, value = TRUE) )

#Alternatively
grep("^South", levels(gapminder$country), value = TRUE)
```

----
# Find countries that have "Ga" at the beginning of either word (not string)

```{r lab2}
grep("\\bGa", levels(gapminder$country), value = TRUE)
```

----
# Write function
* Keep only alpha-numeric characters
* Remove all spaces
* Convert to lower case

```{r lab3}

cleanString <- function(x) {
	string <- gsub("[^[:alnum:]]", "", x)
	string <- tolower(string)
return(string)
}
cleanString("Coeur d'Alene")
```

----
```{r lab4}
gapminder$country[1555]
cleanString(gapminder$country)[1555]
```
