---
title: "Homework 2 Solutions"
output:
  pdf_document:
    highlight: zenburn
---

1. Load data

```{r}
setwd("/Users/Daniel/Dropbox/Teaching/CourseR/data/")
d <- read.csv("CollegeScorecard.csv", na = c("NULL", "PrivacySuppressed"))
```
2. Process data

```{r}
d <- d[ ,c("INSTNM", "STABBR", "SAT_AVG_ALL", "GRAD_DEBT_MDN_SUPP", 
	       "md_earn_wne_p10")]
names(d) <- c("Institution", "State", "SAT", "Debt", "Earnings")

d$SAT_c <- d$SAT - mean(d$SAT, na.rm = TRUE)
mean(d$SAT_c, na.rm = TRUE)
d$Debt_c <- d$Debt - mean(d$Debt, na.rm = TRUE)
mean(d$Debt_c, na.rm = TRUE)
```

3. Create plots

```{r, fig.height = 3}
pairs(d[ ,3:5])
hist(d[ ,3])
hist(d[ ,4])
hist(d[ ,5])
plot(density(d[ ,3], na.rm = TRUE))
plot(density(d[ ,4], na.rm = TRUE))
plot(density(d[ ,5], na.rm = TRUE))
```
\newpage

4. Fit preliminary models

```{r}
mA <- lm(Earnings ~ SAT, data = d)
summary(mA)
mB <- lm(Earnings ~ Debt, data = d)
summary(mB)
mC <- lm(SAT ~ Debt, data = d)
summary(mC)
```

5. Plot regression models

The results of the coefficents are pretty straightforward to interpret. They 
  represent the expected change in the outcome given a 1 unit increase in the 
  predictor. Note that the coefficient for `Debt` is very small, because it 
  represents the expected change in the outcome given a one dollar increase in
  debt. The intercepts range from difficult to interpret, to impossible. For
  example, the SAT scale does not go below 200, yet the intercept in Model A 
  represents the expected earnings for schools with an average SAT score of 0.
  After centering the variables and refitting the models (code below), the
  intercepts represent the expected value of the outcome when the school has an
  average level of the predictor.


```{r, fig.height = 3}
plot(d$Earnings ~ d$SAT)
abline(coef(mA)[1], coef(mA)[2], col = "blue", lwd = 3, lty = 2)
plot(d$Earnings ~ d$Debt)
abline(coef(mB)[1], coef(mB)[2], col = "blue", lwd = 3, lty = 2)
plot(d$SAT ~ d$Debt)
abline(coef(mC)[1], coef(mC)[2], col = "blue", lwd = 3, lty = 2)
```

\newpage

```{r}
mA <- lm(Earnings ~ SAT_c, data = d)
summary(mA)
mB <- lm(Earnings ~ Debt_c, data = d)
summary(mB)
mC <- lm(SAT_c ~ Debt_c, data = d)
summary(mC)
```

6. Fit the multiple regression model

```{r}
mr <- lm(Earnings ~ SAT_c + Debt_c, data = d)
summary(mr)
```

*Extra Credit.* Predictor-residual plot

To compute the predictor residual plot, first predict SAT scores for each
  observation from their Debt.

```{r}
predSAT <- coef(mC)[1] + coef(mC)[2]*d$Debt_c
```
Next compute the residual - i.e., the difference between the predicted and 
  observed SAT score.

```{r}
resSAT <- d$SAT - predSAT
```
Finally, plot the relation between the residualized SAT variable and Earnings. 
  Note that I've added a few additional argument to the plot to put an overall
  title and label the x and y axes.

```{r}
plot(resSAT, d$Earnings, 
	main = "Predictor Residual Plot", 
	xlab = "Residualized SAT (relative to Debt)",
	ylab = "Earnings")
```

The plot above is the relation between `Earnings` and `SAT_c` while controlling
  for (i.e., residualzing for) `Debt_c`. To actually plot the residualzed line,
  you'll need to fit one additional model.

```{r}
resLine <- lm(d$Earnings ~ resSAT)
```
Then you can just use `abline()` like normal

```{r, echo = FALSE}
plot(resSAT, d$Earnings, 
	main = "Predictor Residual Plot", 
	xlab = "Residualized SAT (relative to Debt)",
	ylab = "Earnings")
abline(coef(resLine)[1], coef(resLine)[2], col = "red", lwd = 3, lty = 2)
```

```{r, eval = FALSE}
abline(coef(resLine)[1], coef(resLine)[2], col = "red", lwd = 3, lty = 2)
```


