---
title: "Homework 5 Solutions"
output:
  pdf_document:
    highlight: zenburn
---

```{r echo = FALSE, message = FALSE}
library(openintro)
data(countyComplete)
data(births)
d <- countyComplete
d2 <- births
```

Note for there are multiple ways to complete this homework. The below 
  represent the methods I used (today, as I write this), and alternative methods are certainly possible.


1. Load the *countyComplete* dataset from the *openintro* package. 
	+ Use a `for()` loop to calculate the grand mean for the following variables: `r names(d)[c(10:17, 21, 22, 25:29)]`. 
	+ Calculate the same values using an `apply` family loop.
	+ Conduct a test (i.e., not just visual comparison) to ensure the same means were produced by both methods.

```{r item1}
library(openintro)
data(countyComplete)
d <- countyComplete

varLocs <- c(10:17, 21, 22, 25:29)

# for loop
means <- rep(NA, length(varLocs))
for(i in seq_along(means)) {
	means[i] <- mean(d[ ,varLocs[i]], na.rm = TRUE)
}

round(means, 2)

# apply family loop
means2 <- apply(d[ ,varLocs], 2, mean, na.rm = TRUE)
round(means2, 2)

table(means == means2)
```

2. Calculate the *difference* between the mean for each of the preceding
  variables and the mean for each state. Do so using only `for()` loops, and then again using only the apply family of loops.

```{r item2}

## For loop method

stateMeanDiff <- matrix(#	 NROWS						NCOLS
					rep(NA, length(unique(d$state)) * length(varLocs)),
				ncol = length(varLocs)
				)

s <- split(d[ ,varLocs], d$state)

for(i in seq_along(s)) {
	for(j in 1:ncol(s[[1]])) {
		stateMeanDiff[i, j] <- means[j] - mean(s[[i]][ ,j], na.rm = TRUE)
	}
}

# Make it prettier
dimnames(stateMeanDiff) <- list(names(s), names(s[[1]]))

# Show first six rows and columns
stateMeanDiff[1:6, 1:6] 

## apply family method
stateMeanDiff2 <- sapply(seq_along(varLocs), function(i) {
	means[i] - tapply(d[ ,varLocs[i]], d[ ,1], mean, na.rm = TRUE)
})

#Give column names
colnames(stateMeanDiff2) <- names(d[ ,varLocs])
stateMeanDiff2[1:6, 1:6]
table(stateMeanDiff == stateMeanDiff2)
```

3. Load the *births* dataset. Use `by()` **or** `tapply()` to calculate the
  mean birth weight by the following: `r names(d2)[c(3, 4, 8, 9)]`. Calculate the same means using `aggregate()`
  	+ Use the output to report the following:
  		- Mean birth weights across weeks for full term non-smokers.
  		- Mean birth weights across weeks for premature babies from mothers
  		  who smoke.
  		- Mean birth weights for weeks 38-40 for full term males across smokers
  		- Mean birth weights across weeks for female babies of nonsmoker
  		  mothers, between premie and full term.


```{r item3}
data(births)
d2 <- births

# by method

weightMeans <- by(d2$weight, 
				as.list(d2[ ,c(3, 4, 8, 9)]),
				mean, na.rm = TRUE)
weightMeans[ ,"full term", ,"nonsmoker"]
weightMeans[ ,"premie", ,"smoker"]
weightMeans[10:14, "full term", "male", ]
weightMeans[ , , "female","nonsmoker"]

# aggregate method

weightMeans2 <- aggregate(weight ~ weeks + premature + sexBaby + smoke, 
	data = d2, mean, na.rm = TRUE)
subset(weightMeans2, premature == "full term" & smoke == "nonsmoker")
subset(weightMeans2, premature == "premie" & smoke == "smoker")
subset(weightMeans2, weeks %in% 38:42 & premature == "full term" & 
	sexBaby == "male")
subset(weightMeans2, sexBaby == "female" & smoke == "nonsmoker")
```